---
title: "MLM Project Part 4"
author: "Bianca Brusco, Clare Clingain, Kaushik Mohan, & Frankie Wunschel"
date: "4/26/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(lme4)
library(lattice)
library(lmerTest)
library(ggplot2)
```

#Part 4: Kaushik

## Reload the data and make person-period file

```{r part4 person-period}
#re-read data
classroom <- read.csv("classroom.csv")
classroom2 <- na.omit(classroom)
#new variables
classroom2 <- classroom2 %>% mutate(math0 = mathkind) %>% mutate(math1 = mathkind+mathgain)
#reshape the data
class_pp <- reshape(classroom2, varying = c("math0", "math1"), v.names = "math", timevar = "year",
times = c(0, 1), direction = "long")
```

## Baseline model: unconditional growth model

$$MATH_{tijk} = b_0 + \delta_{0ijk} + (b_1 + \zeta_{1k} )TIME_{tijk} + \zeta_{0k} + \epsilon_{tijk} $$

where $t$ represents occasion (in this case, year/grade), $i$ represents students, $j$ represents classrooms and $k$ represents schools. $\delta_{0ijk} \sim N(0,\sigma^2_{\delta_0})$, $\zeta_{0k} \sim N(0,\sigma^2_{\zeta_0})$, $\zeta_{1k} \sim N(0,\sigma^2_{\zeta_1})$ and $\epsilon_{ijk} \sim N(0,\sigma^2_{\epsilon})$ all independent of each other except for $\zeta_{0k}$ and $\zeta_{1k}$ having a correlation $\rho_{\zeta_{0}\zeta_{1}}$.

```{r part4 UGM}
ugm <- lmer(math ~ year + (year|schoolid) + (1|classid:childid), data=class_pp)
summary(ugm)

```

##  Add student, classroom and school level fixed effects
$$MATH_{tijk} = b_0 + \delta_{0ijk} + (b_1 + \zeta_{1k} )TIME_{tijk} + b_2SEX_{ijk} + b_3SES_{ijk} + b_4MINORITY_{ijk} + b_5YEARSTEA_{jk} + \\ b_6MATHKNOW_{jk} + b_7MATHPREP_{jk} + b_8HOUSEPOV_{k} + \epsilon_{tijk} $$

where $t$ represents occasion (in this case, year/grade), $i$ represents students, $j$ represents classrooms and $k$ represents schools. $\delta_{0ijk} \sim N(0,\sigma^2_{\delta_0})$, $\zeta_{0k} \sim N(0,\sigma^2_{\zeta_0})$, $\zeta_{1k} \sim N(0,\sigma^2_{\zeta_1})$ and $\epsilon_{ijk} \sim N(0,\sigma^2_{\epsilon})$ all independent of each other except for $\zeta_{0k}$ and $\zeta_{1k}$ having a correlation $\rho_{\zeta_{0}\zeta_{1}}$.

```{r part4 add class school}
fit2 <- lmer(math ~ year + sex + ses + minority + yearstea + mathknow + mathprep + housepov + (year|schoolid) + (1|classid:childid), data=class_pp)
summary(fit2)

```

## For year==0:
*what percent of between school differences were explained as you go from the baseline to the second model?
*what percent of between child differences were explained as you go from the baseline to the second model?

## For year==1:
*what percent of between school differences were explained as you go from the baseline to the second model?
*what percent of between child differences were explained as you go from the baseline to the second model?

## Based on significance, 
*  what factors seem useful in describing ("explaining") differences between student outcomes? 
*    Point out the direction of the effect.

## Add random slope for SES

```{r part4 rs ses}

```

## *is the estimated s.d. (square root of variance) of the random slope associated with SES large enough 
*  so that a value +/- 1 s.d. is sufficient to "cancel" (or flip the sign) the fixed effect for this predictor?

## Residuals and Q-Q Plot
*Is there any reason to question the normality assumption?
```{r part4 residuals}

```

## BLUPs for all 4 random effects & Scatter plots

generate an all pairs scatter plot matrix (4x4) of these 
* note whether or not you identify any concerns from these scatterplots.
```{r part4 BLUPs}

```

