---
title: "Part3-B"
author: "Bianca Brusco"
date: "4/28/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
require(lme4)
require(lattice)
require(lmerTest)
library(ggplot2); vanillaR <- F
classroom <- read.csv("classroom.csv")
```


# Part 3: Bianca

## Create person-period file

```{r person-period file}
#re-read data
classroom <- read.csv("classroom.csv")
classroom2 <- na.omit(classroom)
#new variables
classroom2 <- classroom2 %>% mutate(math0 = mathkind) %>% mutate(math1 = mathkind+mathgain)
#reshape the data
class_pp <- reshape(classroom2, varying = c("math0", "math1"), v.names = "math", timevar = "year",
times = c(0, 1), direction = "long")
```

## Initial longitudinal model

The equation for the model below:

$$Math_{tijk} = b_0 + \zeta_{0k} + b_1*Time_{tijk} + \epsilon_{tijk}$$
where $\zeta_{0k} \sim N(0, \sigma_{zeta0}^2)$ and $\epsilon_{tijk} \sim N(0, \sigma_{\epsilon}^2)$

```{r pp model1}
fit1 <- lmer(math ~ year + (1|schoolid), data = class_pp)
summary(fit1)
```

## Add child-level random intercept

$Math_{tijk} = b_0 + \delta_{0tijk} + \zeta_{0k} + b_1*Time_{tijk} + \epsilon_{tijk}$
where $\delta_{0tijk} \sim N(0, \sigma_{\delta_0}^2), \zeta_{0k} \sim N(0, \sigma_{zeta0}^2)$ and $\epsilon_{tijk} \sim N(0, \sigma_{\epsilon}^2)$ independetly of one another. 


```{r pp model2}
fit2 <- lmer(math ~ year + (1|schoolid/childid), data = class_pp)
summary(fit2)
```

In model 1 the variance $\sigma_{\zeta_0}^2 = 377$ and in model 2  $\sigma_{\zeta_0}^2 = 293.2$. 
In model 1 , the varaince for $\sigma_{\epsilon}^2 = 1288$ and in model 2 $\sigma_{\epsilon_0}^2 = 602.2$. We note that including child-level variation leads to a decrease in the variance of both the random effects. 

## Compute Pseudo-R^2

Compute a pseudo R^2 relating the between school variation and ignoring between students in the same school. 

We calculate this as :

$$\frac{\sigma_{\zeta_0}^2(M_0) - \sigma_{\zeta_0}^2(M_1)}{\sigma_{\zeta_0}^2(M_0)} = \frac{377-293.2}{377} = 0.22$$
The between-school variance is reduced by 22% with the introduction of student random effect?

\textbf{Does the total variation stay about the same?}

```{r pp mods totvariance}
tot_m0 = 337 + 1288
tot_m1 = 722 + 293.2 + 602.2 
paste("Tot variance for model 0 : ", tot_m0)
paste("Tot variance for model 1: ", tot_m1)

```

There is only a slightly decrease in the total variance between Model 0 and Model1. 

## Add a random slope for time trend

```{r pp mod3}
fit3 = lmer(math ~ year + (1 + year|| schoolid) + (1|childid), data = class_pp)
summary(fit3)
```


## Generate the BLUPs for this model (fit 3)
Examine then whether the independence between zeta0 and zeta1 is reflected in a scatterplot of these two sets of effects. 

***CHECK: is this what he is asking? i.e. check correlation between zeta0 and zeta1

```{r pp blups}
pp_ranefs <- ranef(fit3)




if (vanillaR) {
plot(pp_ranefs$schoolid[,2],pp_ranefs$schoolid[,1]) 
}else{
ggplot(pp_ranefs$schoolid, aes(x = pp_ranefs$schoolid[,2], y = pp_ranefs$schoolid[,1] )) +
geom_point() + labs(x = "Zeta_{0k} BLOPs", y = "Zeta_{1k} BLOps") + geom_hline( yintercept = 0)
}



#cor(pp_ranefs$schoolid[,2], pp_ranefs$schoolid[,1])
```

Looks a bit like there is **some** negative correlation but not much?

## Heteroscedasticity in the random effects

\textbf{Question:} What are: $V_S(year = 0)$, $V_S(year = 1)$?

```{r}
summary(fit3)
```
The model we are considering is           **(check subscripts pls)

$Math_{tijk} = b_0 + \delta_{0tijk} + \zeta_{0k} + (b_1 + \zeta_{1k})Time_{tijk} + \epsilon_{tijk}$

So we have that (in this model where we are forcing correlation of 0 between slope and intercept):

-  $V_S(year = 0) = \sigma_{\zeta_{0k}}^2 = 85.96$
-  $V_S(year = 1) = \sigma_{\zeta_{0k}}^2 + \sigma_{\zeta_{1k}}^2 = 85.96 + 315.69 = 401.65$

## Run model by year

```{r}
class_year0 = class_pp[class_pp$year == 0,]



fit4 = lmer(math ~ (1 | schoolid), data = class_year0)
summary(fit4)


class_year1 = class_pp[class_pp$year == 1,]
fit5 = lmer(math ~ (1 | schoolid), data = class_year1)
summary(fit5)



```

No, we do no ge the same estimates for the variances by running the model separately. 

## Allow for correlation 

```{r}
fit6 = lmer(math ~ year + (1 + year| schoolid) + (1|childid), data = class_pp)
summary(fit6)

```

Correlation beteween $\zeta_0$ and $\zeta_1$ = -0.53. 
Statistically significant?

So we have that (in this model where we are forcing correlation of 0 between slope and intercept):

-  $V_S(year = 0) = \sigma_{\zeta_{0k}}^2 = 373.5$
-  $V_S(year = 1) = \sigma_{\zeta_{0k}}^2 + \sigma_{\zeta_{1k}}^2 + 2\rho_{01}\sigma_{\zeta_{0k}}\sigma_{\zeta_{1k}}= 373.5 + 112.4 - 2*0.53*\sqrt{373.5}\sqrt{112.4} = 268.7$

These are a lot closer to the school variances that result from fitting the models for the two years separately ( in which we have $\sigma_{zeta}^2$ respectively be 364 for year 0 and 279 for year 1. 