---
title: "MLM Project Part 4"
author: "Bianca Brusco, Clare Clingain, Kaushik Mohan, & Frankie Wunschel"
date: "4/26/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(lme4)
library(lattice)
library(lmerTest)
library(ggplot2)
```

#Part 4: Kaushik

## Reload the data and make person-period file

```{r part4 person-period}
#re-read data
classroom <- read.csv("classroom.csv")
classroom2 <- na.omit(classroom)
#new variables
classroom2 <- classroom2 %>% mutate(math0 = mathkind) %>% mutate(math1 = mathkind+mathgain)
#reshape the data
class_pp <- reshape(classroom2, varying = c("math0", "math1"), v.names = "math", timevar = "year",
times = c(0, 1), direction = "long")
```

## Baseline model: unconditional growth model
mixed math year || schoolid: year, cov(un) || childid:
WRITE THE MODEL EQUATION

```{r part4 UGM}

```

##  Add classroom and school level fixed effects
WRITE THE MODEL EQUATION
```{r part4 add class school}

```

## For year==0:
*what percent of between school differences were explained as you go from the baseline to the second model?
*what percent of between child differences were explained as you go from the baseline to the second model?

## For year==1:
*what percent of between school differences were explained as you go from the baseline to the second model?
*what percent of between child differences were explained as you go from the baseline to the second model?

## Based on significance, 
*  what factors seem useful in describing ("explaining") differences between student outcomes? 
*    Point out the direction of the effect.

## Add random slope for SES

```{r part4 rs ses}

```

## *is the estimated s.d. (square root of variance) of the random slope associated with SES large enough 
*  so that a value +/- 1 s.d. is sufficient to "cancel" (or flip the sign) the fixed effect for this predictor?

## Residuals and Q-Q Plot
*Is there any reason to question the normality assumption?
```{r part4 residuals}

```

## BLUPs for all 4 random effects & Scatter plots

generate an all pairs scatter plot matrix (4x4) of these 
* note whether or not you identify any concerns from these scatterplots.
```{r part4 BLUPs}

```

